---
editor_options:
  chunk_output_type: console
output:
  html_document: default
  pdf_document: default
---
Amr Mohammad Soliman 
========================================================

This data set contains 113,937 loans with 81 variables on each loan,
including loan amount, borrower rate (or interest rate), 
current loan  status, borrower income, borrower employment 
status,and the latest payment information.


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(GGally)
library(dplyr)
library(gridExtra)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
pld <- read.csv("prosperLoanData.csv")
```
# Univariate Plots Section

## BorrowerAPR and BorrowerRate

#### Histogram

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots}
# Plot a histogram of the Borrower Rate
p1 <- ggplot(aes(x = BorrowerRate), data = pld) + geom_histogram() 

# Plot a histogram of the Borrower APR
p2 <- ggplot(aes(x = BorrowerAPR), data = pld) + geom_histogram() 

grid.arrange(p1, p2, ncol = 2)
```



#### Boxplots

```{r echo=FALSE,message=FALSE, warning=FALSE}
# Plot a boxplot of the Borrower APR
p1 <- ggplot(aes(x=1, y=BorrowerAPR), data = pld) +
  geom_boxplot() + 
  coord_cartesian(ylim = c(0,0.3))

# Plot a boxplot of the Borrower Rate
p2 <- ggplot(aes(x = 1, y=BorrowerRate), data = pld) + 
  geom_boxplot() + 
  coord_cartesian(ylim = c(0,0.3))

grid.arrange(p1, p2, ncol = 2)
```


### Display the mean and range of the BorrowerRate

```{r echo=FALSE,message=FALSE, warning=FALSE}
summary(pld$BorrowerRate)
```

### Display the mean and range of the BorrowerAPR

```{r echo=FALSE,message=FALSE, warning=FALSE}
summary(pld$BorrowerAPR)
```
The distribution of BorrowerAPR and BorrowerRate should be\
similar with  slight difference since the APR is always\
higher than the interest rate.

## Employment status

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x = EmploymentStatus), 
       data = subset(pld, !is.na(EmploymentStatus))) +
  geom_histogram(stat = "count") +
  theme(axis.text.x = element_text(angle = 30))
```

```{r echo=FALSE,message=FALSE, warning=FALSE}
by(pld$BorrowerRate, pld$EmploymentStatus, summary)
```
The vast majority of borrowers are Employed/
following by Full-time

## Income Range buckets

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(pld, aes(x = IncomeRange)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))
```

The income ranges show that the vast majority of\
borrowers make less than $75,000 per year.

## Loan Status 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(pld, aes(x = LoanStatus)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))
```

It appears that a large majority of loans are/
either Completed or Curren

## The Term to maturity for each loan

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(pld, aes(x = Term)) +
  geom_histogram() 

summary(pld$Term)
```

The majority of of loans are short term at 36 months. 

## Histogram of Stated Monthly Income

```{r echo=FALSE,message=FALSE, warning=FALSE}

p1 <- ggplot(pld, aes(x = StatedMonthlyIncome),  binwidth = 30) +
  geom_histogram() 

p2 <- ggplot(pld, aes(x = StatedMonthlyIncome),  binwidth = 30) +
  geom_histogram() + 
  scale_x_continuous(limits = c(0, quantile(pld$StatedMonthlyIncome, 0.99)))

p3 <- ggplot(pld, aes(x = StatedMonthlyIncome),  binwidth = 30) +
  geom_histogram() + 
  scale_x_log10()

grid.arrange(p1, p2, p3, ncol = 3)

summary(pld$StatedMonthlyIncome)
```

The histogram plot of the original data is highly right skewed.\
So I used summary function and find that the maximum\
value is 1750003, which is likely to be an error.\
So in the next 2 plots, I removed the top 1% data\
from the original and I used scale_x_log10 to get\
more reasonable plot.

## Histogram of the Debt To Income Ratio

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x = DebtToIncomeRatio),data = subset(pld, !is.na(DebtToIncomeRatio))) +
 geom_histogram() +
scale_x_continuous(limits = c(0,1))

summary(pld$DebtToIncomeRatio)
```

The histogram is right skewed, after filtering the NA values\
and excluding the outliers.After using summary function,\
I find that the maximum value is 10.010\
,which is likely to be an error.

## Histogram of the Available Bank card Credit

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(pld, aes(x = AvailableBankcardCredit)) +
 geom_histogram() +
  scale_x_continuous(limits = c(0, 1e+05))

summary(pld$AvailableBankcardCredit)
```

Most of borrowers have low values of Available Bank card Credit


## BorrowerRate by credit grade

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerRate), data = subset(pld, !is.na(CreditGrade))) + geom_histogram() +
  facet_wrap(~CreditGrade)
```

```{r echo=FALSE,message=FALSE, warning=FALSE}
by(pld$BorrowerRate, pld$CreditGrade, summary)
```

The most of BorrowerRate is (B) Grade

## BorrowerState

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerState), 
       data = subset(pld, !is.na(BorrowerState))) +
  geom_histogram(stat = 'count') +
  theme(axis.text.x=element_text(angle=90, hjust=1, size=6))

```

Most of borrowers are from CA

## Occupation

```{r echo=FALSE,message=FALSE, warning=FALSE}
 ggplot(aes(x = Occupation), data = subset(pld, !is.na(Occupation))) + geom_histogram(stat = 'count') +
  theme(axis.text.x = element_text(angle = 90, hjust=1, size = 6))
```

‘Other’ and ‘Professional’ stood out in the occupation chart. \ 
This shows that many borrowers decided to choose \ 
“other” and “professional” instead of their real occupation.


```{r echo=FALSE,message=FALSE, warning=FALSE}
#Removing Other and professional
occ=subset(pld, pld$Occupation!='Other' & 
               pld$Occupation!='Professional' )

ggplot(aes(x=Occupation), data = occ)+
  geom_bar()+
  ylab('Counts')+
  theme(axis.text.x=element_text(angle=90, hjust=1, size = 6))
```


After removing the 'Others' and 'Professional' data, \ 
the top three occupations counts are 'computer programmer', \ 
'Excutives', and 'Teacher'. 


```{r echo=FALSE}
occ=subset(pld, pld$Occupation!='Other' & 
               pld$Occupation!='Professional' )
ggplot(aes(x=Occupation), data = occ)+
  geom_bar()+
  ylab('Counts')+
  theme(axis.text.x=element_text(angle=90, hjust=1, size = 6))
```

## BorrowRAte by IsBorrowerHomeowner

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerRate), data = subset(pld, !is.na(IsBorrowerHomeowner))) + geom_histogram() +
  facet_wrap(~IsBorrowerHomeowner)

qplot(x = BorrowerRate, data = subset(pld, !is.na(IsBorrowerHomeowner)), 
      geom = 'freqpoly',color = IsBorrowerHomeowner) 

by(pld$BorrowerRate, pld$IsBorrowerHomeowner, summary)
```

The borrowers who aren't homeowners are slighty\
more than the homeowners

## BankCardUtilization

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x = BankcardUtilization), data = pld) + 
  geom_histogram() +
  scale_x_continuous(limits = c(0,1))

summary(pld$BankcardUtilization)

```
Being left skewed makes sense, because most of the \ 
borrowers might have almost maxed out their credit limits

## Rating

```{r echo=FALSE,message=FALSE, warning=FALSE}
p1 <- ggplot(data = subset(pld, !is.na(ProsperRating..Alpha.))) +
  geom_bar(aes(x = ProsperRating..Alpha.))

p2 <- ggplot(data = subset(pld, !is.na(ProsperRating..numeric.))) +
  geom_bar(aes(x = ProsperRating..numeric.))

grid.arrange(p1, p2,ncol = 2)
```

#### Summary of ProsperRating..Alpha.
```{r echo=FALSE,message=FALSE, warning=FALSE}
summary(pld$ProsperRating..Alpha.)
```

#### Summary of ProsperRating..numeric.
```{r echo=FALSE,message=FALSE, warning=FALSE}
summary(pld$ProsperRating..numeric.)
```

#### The mean of ProsperRating..numeric. for ProsperRating..Alpha.
```{r echo=FALSE,message=FALSE, warning=FALSE}
by(pld$ProsperRating..numeric., pld$ProsperRating..Alpha., mean)
```
The most common rating is C(4) follows by B(5),\
A(6) and D(3).The most common of borrowers have\
rating (C) The highest rating (AA) has the lowest\
number of borrowers.


# Univariate Analysis

### What is the structure of your dataset?
There are 113,979 loan lists in the data set with 81 variables\
from 2005 Nov to 2014 Mar. 
```{r, echo=FALSE}
str(pld)
```

### What is/are the main feature(s) of interest in your dataset?
* BorrowerAPR
* BorrowerRate
* BorrowerState
* Employment status
* Income Range buckets
* Loan Status
* The Term
* Stated Monthly Income
* Debt To Income Ratio
* Available Bank card Credit
* Credit grade
* Occupation
* BankCardUtilization

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

* Employment Duration
* Loan Original Amount
* Investors
* Employment Status
* Prosper Rating

### Did you create any new variables from existing variables in the dataset?
No, I didn't

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

No, I didn't perform any operations on the data



# Bivariate Plots Section
## Scatterplot Matrices

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggpairs(pld[,c('LoanOriginalAmount', 'Investors', 'Term',
                'EmploymentStatus', 'ProsperRating..Alpha.',
                'BorrowerRate', 'MonthlyLoanPayment')])
```


MonthlyLoanPayment has 2 stronge relationships with\
LoanOriginalAmount and Term

## The relationship between Loan Original Amount and Number of Investors

```{r echo=FALSE,warning=FALSE}
ggplot( data = pld) + 
  geom_jitter(aes(x = LoanOriginalAmount, y = Investors),alpha = 1/20) +
 
  ggtitle("Loan Original Amount vs Number of Investors") +
  xlab("Loan Original Amount")
```


Larger loans have more investors

## Relationship between loan amount and terms.

```{r echo=FALSE, warning=FALSE}
ggplot(data = pld) + 
  geom_boxplot(aes(x = as.character(Term), y = LoanOriginalAmount)) +
  ggtitle("Term vs Loan Original Amount") +
  xlab("Term") +
  ylab("Loan Original Amount")
```

```{r echo=FALSE, warning=FALSE} 
by(pld$LoanOriginalAmount, pld$Term, summary)
```


The loan amonts are getting larger with longer terms. 

## Employment status and loan amount

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = EmploymentStatus, y = LoanOriginalAmount), data = pld) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, quantile(pld$LoanOriginalAmount, 0.99))) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Employed people requested more loans and not employed people\
wanted more loan than part-time employed people

## Prosper Rating against borrower rate

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = ProsperRating..Alpha., y = BorrowerRate), data = pld) + 
  geom_boxplot()
```

As we can see The better rating belongs to lower borrower rate. 

## Relationship of term and borrower rate

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = factor(Term), y = BorrowerRate), data = pld) +
 geom_boxplot() + 
  coord_cartesian(ylim = c(0,0.27))
```

The median of 60 month term is the highest one,\
while the 12 month has the minimum one.

## MonthlyLoanPayment vs. Occupations

```{r echo=FALSE}
ggplot(aes(y=MonthlyLoanPayment, x=Occupation), data = pld)+
  stat_summary(fun.y=mean, geom='bar')+
  theme(axis.text.x=element_text(angle=90, hjust=1,size=6))+
  geom_bar(data = subset(pld,
                    pld$Occupation=='Judge'|
                    pld$Occupation=='Doctor'|
                    pld$Occupation=='Pharmacist'), 
           aes(y=MonthlyLoanPayment, x=Occupation), 
           stat = 'summary', 
           fun.y=mean, 
           fill="#cf4c35")+
  ylab('Monthly Payment ($)')
  
```

The top 3 occupations have high monthly loan payment are Judge,\
Doctor, and Pharmacist.The groups that have lower monthly loan\ 
payment are students.



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

* The number of investors is increasing with higher prosper score.
* The better rating belongs to lower borrower rate.
* The loan amonts are getting larger with longer terms. 
* Employed people requested more loans and not employed people
* The top 3 occupations have high monthly loan payment are\
Judge, Doctor,and Pharmacist.The monthly loan payment should\
be positively correlated with the salary.The top 3 listed\
occupations are all best-paying jobs according to\ [CNBC](http://www.cnbc.com/2016/07/26/the-25-highest-paying\
-jobs-in-america.\html), and the groups that have lower monthly loan\ 
payment are students.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I noticed that employed people are more likely to\
loan than others. Also debt to income ratio for\
rating AA is the lowest one.

### What was the strongest relationship
you found?

MonthlyLoanPayment with Term (0.932)


# Multivariate Plots Section

```{r echo=FALSE,warning=FALSE, message=FALSE}
# create new datetime variable for the listing created date
pld$ListingDateTime <- strptime(x = as.character(pld$ListingCreationDate),
                                format = "%Y-%m-%d")

pld$Month <- as.Date(cut(pld$ListingDateTime,
  breaks = "month"))

pld$BankcardUtilization.bucket <- cut(pld$BankcardUtilization, 
                                           c(0, 0.25, 0.5, 0.75, 1.0, 5.0))
```
## Scatterplot Matrices
```{r echo=FALSE, warning=FALSE, message=FALSE,Multivariate_Plots}
ggpairs(pld[,c('Month','BankcardUtilization.bucket',
               'BankcardUtilization','IncomeRange', 
               'IsBorrowerHomeowner','BorrowerRate', 
               'CreditScoreRangeLower')])
```

## Borrower Rate vs Income Range vs Is Borrower Homeowner

```{r echo=FALSE,warning=FALSE, message=FALSE}
ggplot(aes(x = as.factor(IncomeRange), y = BorrowerRate), 
       data = subset(pld, 
                     IncomeRange != 'Not employed' & 
                       IncomeRange != 'Not displayed'& 
                       IncomeRange != '$0')) +
  geom_boxplot(aes(fill = IsBorrowerHomeowner)) +
  ggtitle("Income Range vs Borrower Rate (by Home Owner)") +
  guides(fill = guide_legend(title = "Is Borrower Homeowner")) +
  xlab("Income Range") +
  ylab("Borrower Rate")  +
   theme(text = element_text(size=12), 
        axis.text.x = element_text(angle=90, vjust=1))
```

As we can see borrowers with higher income, who are home owners\
have lower borrower rates. 

## Credit Scores vs Borrower rate vs Bankcard Utilization 
```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(data = subset(pld, CreditScoreRangeLower > 400 & 
                       BankcardUtilization < 5.0 & 
                       !is.na(BankcardUtilization)), 
       aes(x = factor(CreditScoreRangeLower), 
           y = BorrowerRate, 
           fill = BankcardUtilization.bucket)) + 
  geom_boxplot() +
  theme(legend.position = 'bottom')
```

In general, a higher bankcard utilization rate corresponds\
with lower credit scores and higher interest rates.

## Interest Rate vs. Time vs. Homeowner

```{r echo=FALSE,warning=FALSE, message=FALSE}
ggplot(aes(x=Month, y=BorrowerRate, color=as.factor(IsBorrowerHomeowner)),
       data = subset(pld, !is.na(pld$IsBorrowerHomeowner)))+
  stat_summary(fun.y = mean, geom = 'point') + 
  geom_smooth(method = "lm", se = FALSE,size=1) +
  scale_color_discrete(guide = guide_legend(title = "Homeowner"),
                       labels=c('No','Yes'))+
  xlab('Time')+
  ylab('Borrower Rate (%)')
```

The interest rate for homeowners are generally lower\
than non-homeowners

## Relationship between Delinquencies and Employment Status

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x = DelinquenciesLast7Years, y = AmountDelinquent,
                      color = as.factor(EmploymentStatus)),
       data = subset(pld, AmountDelinquent > 0 & 
                            EmploymentStatus != "" )) +
  geom_point(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=2) +
  xlim(0, quantile(pld$DelinquenciesLast7Years, 0.99, na.rm = T)) +
  ylim(0, quantile(pld$AmountDelinquent, 0.99, na.rm = T)) +
  scale_color_discrete(guide = guide_legend(title = "EmploymentStatus")) +
ggtitle('Relationship between Delinquencies and Employment Status')
```

From the above we can see relationship between the\
amount borrowers were delinquent and the number of\
delinquencies they've had over the last 7 years then\
separated that by employment status.It is obvious that\
$Self-employed$ then $Other$ are the maximum. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
* Homeowners usually have lower risk of making defaulted loans
  since bank can use their house as collateral.
  
* A higher bankcard utilization rate corresponds with lower credit
  scores and higher interest rates.

### Were there any interesting or surprising interactions between features?
* I've observed that employed people are more likely to borrow money.


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
There was no need for any models
------

# Final Plots and Summary


### Plot One

##### Histogram of the Debt To Income Ratio

```{r echo=FALSE,message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x = DebtToIncomeRatio),
       data = subset(pld, !is.na(DebtToIncomeRatio))) +
 geom_histogram(binwidth = 0.01 ,color = I('black'), fill = I('#099DD9')) +
scale_x_continuous(limits = c(0,1)) +
ggtitle("Debt To Income Ratio") +
  xlab("Ratio")  
```

#### Description One

The histogram is right skewed, after filtering the NA values\
and excluding the outliers.After using summary function,I find\
that the maximum value is 10.010,which is likely to be an error\
.The highest ratio (peak) is 0.018. 


### Plot Two

##### MonthlyLoanPayment vs. Occupations

```{r echo=FALSE, Plot_Two}
ggplot(aes(y=MonthlyLoanPayment, x=Occupation), data = pld)+
  stat_summary(fun.y=mean, geom='bar')+
  theme(axis.text.x=element_text(angle=90, hjust=1,size=6))+
  geom_bar(data = subset(pld,
                    pld$Occupation=='Judge'|
                    pld$Occupation=='Doctor'|
                    pld$Occupation=='Pharmacist'), 
           aes(y=MonthlyLoanPayment, x=Occupation), 
           stat = 'summary', 
           fun.y=mean, 
           fill="#cf4c35")+
  geom_bar(data = subset(pld,
                    pld$Occupation=='Student - College Freshman'|
                    pld$Occupation=='Student - College Graduate Student'|
                    pld$Occupation=='Student - College Junior'|
                    pld$Occupation=='Student - College Senior'|
                    pld$Occupation=='Student - College Graduate Student'|
                    pld$Occupation=='Student - College Sophomore'|
                    pld$Occupation=='Student - Community College'|
                    pld$Occupation=='Student - Technical School'  
                      ), 
           aes(y=MonthlyLoanPayment, x=Occupation), 
           stat = 'summary', 
           fun.y=mean, 
           fill="#CFBC5F")+
  
  ylab('Monthly Payment [USD]')
```

#### Description Two

The top 3 occupations have high monthly loan payment are $Judge$, $Doctor$,\
and $Pharmacist$.The monthly loan payment should be positively correlated\
with the salary.The top 3 listed occupations are all best-paying jobs\
according to [CNBC](http://www.cnbc.com/2016/07/26/the-25-highest-paying\
-jobs-in-america.\html), and the groups that have lower monthly loan\ 
payment are students.

### Plot Three

##### Relationship between Delinquencies and Employment Status

```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Three}
ggplot(aes(x = DelinquenciesLast7Years, y = AmountDelinquent,
                      color = as.factor(EmploymentStatus)),
       data = subset(pld, AmountDelinquent > 0 & 
                            EmploymentStatus != "" )) +
  geom_point(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0, quantile(pld$DelinquenciesLast7Years, 0.99, na.rm = T)) +
  ylim(0, quantile(pld$AmountDelinquent, 0.99, na.rm = T)) +
  scale_color_discrete(guide = guide_legend(title = "EmploymentStatus")) +
ggtitle('Relationship between Delinquencies and Employment Status') +
  xlab('DelinquenciesLast7Years [USD]') +
  ylab('AmountDelinquent [USD]')
```

#### Description Three
From the above we can see relationship between the\
amount borrowers were delinquent and the number of\
delinquencies they've had over the last 7 years then\
separated that by employment status. Is obvious that\
Employed and Full time are the maximum.\

------

# Reflection

- This was somehow challenging but a good way to have a firm\
grasp on doing exploratory data analysis with R. much\
more difficult than I thought it would be.Specially it was hard\
to get started without any financial knowledge., but in the end,\
it was incredibly rewarding.


- A simple way to show relationships among multi variables\
is to add colors  to the plot. Using different colors to\
represent factorial variables. The  point size of the scatter\
plot can also be used to represent numerical  variables with\
larger size associated with larger numerical values.

- In the future, I think the analysis could be expanded to\
try to predict interestrates on new loans.I think there is\
enough information from the propser data that when combined\
with general market yield curves, it could come close to\
providing some prediction on where new loans would be given\
a set of borrower characteristics.

- The followings are some of the interesting features which\
I observed during the exploratory analysis:

  - Most of borrowers Income ranges from 25,000 - 74,999.
  - There are 3 loan terms 12, 36 and 60 months and the most
    popular one is 36 month.
  - Borrower who home owner usually receive bigger loan than others.
  - Employed person are more likley to take loan than others.
  - Homeowners usually have lower risk of making defaulted loans
    since bank can use their house as collateral.

  
# References
- http://www.cookbook-r.com
- https://www.statmethods.net/index.html
- https://www.r-bloggers.com
- http://www.r-tutor.com
- https://jianru-shi.github.io/Prosper_Loan_EDA/Prosper_Loan_EDA.html

